---
title: "Quantile Regression with `quantreg`"
author: "N. Gennaro - F. Pontarin"
date: "28 giugno 2018"
header-includes:
  - \usepackage{bm}
output:
  beamer_presentation:
    keep_tex: true
    latex_engine: xelatex
    toc: yes
  pdf_document:
    latex_engine: xelatex
    toc: yes
  ioslides_presentation:
    toc: yes
  slidy_presentation:
    toc: yes
---

```{r setup, include=FALSE}
library(MASS)
library(quantreg)
library(rgl)
library(rglwidget)
library(grid)
library(gridBase)
library(ggplot2)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})


hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(if (abs(lines[1])>1 | lines[1]<0) more else NULL,
           x[lines],
           if (length(x)>lines[abs(length(lines))]) more else NULL
          )
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
}) 

knitr::knit_hooks$set(webgl = hook_webgl)

```

# A bit of Theory

## Expanding the usual approach

The objective of linear regression analysis is the estimation of the *mean*.

Quantile Regression aims at estimating specific *quantiles* (e.g. median)


## Expanding the usual approach
```{r echo=FALSE, fi}
data(engel)
plot(foodexp ~ income, data = engel,
main = "Engel data", xlab ="Income", ylab="Food Expenditure", pch='*', pin=c(2,1))
taus <- c(.05, .25, .50, .75, .95)

rqs <- as.list(taus)

for(i in seq(along = taus)) {
rqs[[i]] <- rq(foodexp ~ income, tau = taus[i], data = engel)
  if (taus[i]==0.50){
    abline(rqs[[i]],col=i+1,lty=1,lwd=3)
  } else {
    abline(rqs[[i]],col=i+1,lty=3,lwd=3)
  }
}

lr<-lm(foodexp ~ income, data = engel)
abline(lr,col=8,lty=2,lwd=4)
#lines(engel$income, fitted(lr), col = 8, lwd=1.5, lty=2)
legend("bottomright",1, legend= c(paste(taus, " quantile"),"mean"),title="Est. quantities",
  col = c((2:(length(taus)+1)),8), lty=c(3,3,1,3,3,2), lwd=c(rep(3,length(taus)),4))
```

## Expanding the usual approach

- Linear regression is supported by a number of assumptions about the predictor variables, the response variables and their relationship (error independence, homoscedasticity,... )

The structure underlying this type problems allow them to be transformed into linear algebra equivalents

- On the other hand, Quantile regression makes no particular assumption on the distribution of the response nor about its variance

This leads instead to linear programming problems

## Expanding the usual approach

| Linear Regression               | Quantile Regression             |
| :-----------------------------: | :-----------------------------: |
| Predicts the mean               | Predicts conditional quantiles  |
| Applies when n is small         | Needs sufficient data           | 
| Often assumes normality         | Is distribution agnostic        | 
| Is sensitive to outliers        | Is robust to response outliers  |
| Is computationally inexpensive  | Is computationally intensive    |


Table: A quick comparison between the two regressions


## Under the hood

Called $\tau$ the objective quantile of a random variable $Y$, Quantile regression minimizes a sum that gives asymmetric penalties:

- $(\tau -1) e_i$ for underestimates

- $\tau e_i$ for overestimates

This is condensed in what's defined as the *loss function*

$$
\rho_\tau(y) =  (\tau- 1_{ \{y<0\} }) y
$$

## Under the hood

The wanted quantile $u$ can be calculated by minimizing the expected loss for $Y-u$ that is

$$
E [\rho_\tau(Y-u)] = (\tau-1)\int_{-\infty}^u {(y-u) dF_Y(y)} + \tau \int_u^{+\infty} (y-u) dF_Y(y)
$$

The concept is generalized to allow its use with distributions coming from observed samples

## A visual hint

```{r echo=FALSE, results='hide',message=FALSE}
library(animation) 
library(plyr)

oopt = ani.options(nmax = 150, interval=0.08333, ani.width=800, ani.height=500 )

xmin<- -3
xmax<- 3

xx<- seq(xmin,xmax,0.05)

pos<-c()
vals<-c()
areas<-c()

myplot <-function(xmin,xmax,h,xx,quant,tau,app) {
  
  # plot 1 con la distribuzione e l'effetto della tau sulla distribuzione
  ymin<- -0.1
  ymax<- 1
  discr<-qnorm(quant)
  plot(xx,dnorm(xx), xlim=c(xmin,xmax), ylim=c(ymin,ymax), xlab="u", ylab="density", type="l", lwd=2, col=1)
  lines(c(xmin,discr),c(1-tau,1-tau), lwd=2, col=2)
  lines(c(discr,xmax),c(tau,tau), lwd=2, col=3)
  area1 <- (1-tau)*quant
  area2 <- (tau)*(1-quant)
  text(-1.5,0.9,paste("area =", round(area1,digits=3)),col=2,cex=1.5)
  text(1.5,0.9,paste("area =",round(area2,digits=3)),col=3,cex=1.5)
  
  absdiff<- abs(area2-area1)
  text(0,0.95,paste("absdiff =",round(absdiff,digits=3)),col=1,cex=2)
  
  text((xmin+discr)/2,1-tau+0.06,bquote(1- tau == .(round(1-tau,digits=3))),col=2,cex=1.25)
  text((xmax+discr)/2,tau+0.06,bquote(tau == .(round(tau,digits=3))),col=3,cex=1.25)
  
  text(discr,-0.05,paste(round(discr, digits=3)),col=1,cex=1.5)
  
  curve(dnorm(x)*(x<=discr)*(1-tau), xlim=c(xmin,discr), ylim=c(ymin,ymax), type="l", lwd=2, col=2,lty=2, add=TRUE)
  curve(dnorm(x)*(x>=discr)*tau, xlim=c(discr,xmax), ylim=c(ymin,ymax), type="l", lwd=2, col=3,lty=2, add=TRUE)
  
  segments(discr,0,discr,tau,lty=2,col=1,lwd=1)
  abline(h=0,lwd=1)
  return (c(discr,absdiff))
  
}


myplot2 <-function(pos,vals,areas) {
  ymin<- -0.1
  ymax<- 1.75
  xmin<- -3
  xmax<- 3
  
  #plot 2 con i valori della funzione di costo e differenza tra le aree
  plot(pos,vals, xlim=c(xmin,xmax), ylim=c(ymin,ymax), xlab="u", ylab="value", type="l", lwd=2, col=2)
  lines(pos,areas, xlim=c(xmin,xmax), ylim=c(ymin,ymax), type="l", lwd=2, col=1)
  legend(x= "topright", y=1, legend=c("Abs diff of the areas", expression("Loss function " ~ E* "[" ~rho[tau] (Y-u)~"]") ),
       col=c(2, 1), lwd=c(2,2), lty=c(1,1), cex=1.5)
  abline(h=0,lwd=1)
}


saveVideo({
  
for (h in 1:10) {
  hh<-0
  quant<-0.5
  tau<- 0.5
    
  par(mfrow=c(2,1),mar=c(5, 4, 0, 1.5) + 0.1, cex.axis=1.5, cex.lab=1.5)
  myplot(xmin,xmax,hh,xx,quant,tau,FALSE)
  myplot2(0,0,0)

  ani.pause()
}
  
for (h in 1:70) {
  if (h<=41) {
    hh<-h
  } else {
    hh<-41
  }
  quant<-0.5
  tau<- 0.5+(0.00625*(hh-1))
    
  par(mfrow=c(2,1),mar=c(5, 4, 0, 1.5) + 0.1, cex.axis=1.5, cex.lab=1.5)
  myplot(xmin,xmax,hh,xx,quant,tau,FALSE)
  myplot2(0,0,0)

  ani.pause()
}

for (h in 1:100) {
  if (h<=79) {
    hh<-h
  } else {
    hh<-79
  }
  quant <- 0.0125+0.0125*(hh-1)
  tau<- 0.75
  
  par(mfrow=c(2,1),mar=c(5, 4, 0, 1.5) + 0.1, cex.axis=1.5, cex.lab=1.5)
  vv<-myplot(xmin,xmax,hh,xx,quant,tau,TRUE)
  dd<-vv[1]
  pos<-append(pos,dd)
  vals<-append(vals,vv[2])

  curarea<- (tau-1)*(-dnorm(dd) -dd*pnorm(dd)) + tau*(dnorm(-dd)-dd*pnorm(-dd))
  areas<-append(areas,curarea)
  myplot2(pos,vals,areas)
  
  idx<-which.min(vals)

  segments(pos[idx],0,pos[idx],areas[idx],col=3,lwd=2)
  segments(pos[idx],0,pos[idx],vals[idx],col=3,lwd=2)
  points(pos[idx],areas[idx],col=2,lwd=4)
  points(pos[idx],vals[idx],col=2,lwd=4)
  text(pos[idx],-0.07,paste(round(pos[idx], digits=3)),col=1,cex=1.5)
  text(pos[idx],areas[idx]+0.2,paste(round(areas[idx], digits=3)),col=1,cex=1.5)
  
  ani.pause()
}

}, video.name = "quant.mp4")
ani.options(oopt)
```
<video controls loop autoplay>
  <source src= "quant.mp4" type="video/mp4">
</video>


# The package `quantreg`

# The dataset

# Some Examples


##

to perform a Quantile regression one can call:

```{ size="small"}
rq(formula, tau=.5, data, subset, weights, na.action,
   method="br", model = TRUE, contrasts, ...)
```

So for example:
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, size="small"}
taus <- c(.05,.1,.25,.50,.75,.90,.95)
for( i in 1:length(taus)){
  current.rq<- rq( median_house_value ~ median_income,
             data=house,tau=taus[i])
  abline(current.rq)
}
```

##

```{r echo=FALSE, message=FALSE, warning=FALSE}

house <- read.csv("house_from_gitsklearn.csv")

par(mar=c(4,5,0,1))
plot(house$median_income, house$median_house_value, cex=.25, cex.lab=1,cex.axis=1,
     lheight=0.5,
     xlab="Median Income", ylab="Median House Value")

taus <- c(.05,.1,.25,.75,.90,.95)
for( i in 1:length(taus)){
  abline(rq( median_house_value ~ median_income,
             data=house,tau=taus[i]), col="green3", lwd=2)
}

median.rq<-rq( median_house_value ~ median_income,
             data=house,tau=0.5)
abline(median.rq, col=4, lwd=2)


fit <- lm( median_house_value ~ median_income + (median_income*median_income), data=house )
abline(fit,col=2, lwd=2, lty =2 )

```

##

With the result given by `rq` one can print the summary with the function `summary.rqs`

```{ size="small"}
summary.rqs(object, se = NULL, covariance=FALSE,
            hs = TRUE,  U = NULL, gamma = 0.7, ...)
```

the coefficients for each requested quantiles are shown

```{r warning=FALSE, size="small",  output.lines=10}
population.rq<- rq(population ~ households,
                   data=house, tau=taus)
population.summary<-summary.rqs(population.rq, ci='boot')
population.summary
```


##

One can plot the summary with the following command:


```{ size="small"}
plot.summary.rs(x, parm = NULL, level = 0.9, ols = TRUE,
  mfrow = NULL, mar = NULL, ylim = NULL, main = NULL,
  col = gray(c(0, 0.75)), border = NULL, lcol = 2,
  lty = 1:2, cex = 0.5, pch = 20, type = "b", xlab = "",
  ylab = "", ...)
```


The coefficients for each requested quantiles are plotted together with their confidence intervals.


##
```{r size="small"}
plot.summary.rqs(population.summary)
```

##

`rqss` fits additive quantile regression models with possible univariate and/or bivariate
nonparametric terms.


```
rqss(formula, tau = 0.5, data = parent.frame(), weights,
     na.action, method = "sfn", lambda = NULL,
     contrasts = NULL, ztol = 1e-5, control, ...)
```


```{r warning=FALSE, message=FALSE, size="small"}
fhouse<-house
fhouse$longitude <- round(fhouse$longitude * 100)
fhouse$latitude <- round(fhouse$latitude * 100)
fit <- rqss(median_house_value ~ qss(cbind(longitude,latitude),
                                     lambda = 8),
            data = fhouse)
```

##

The resulting fit can be plotted with the function `plot.rqss`

```{r eval=F, size="tiny"}

plot.new() 

gl <- grid.layout(nrow=1, ncol=2)
vp.1 <- viewport(layout.pos.col=1, layout.pos.row=1) 
vp.2 <- viewport(layout.pos.col=2, layout.pos.row=1) 

pushViewport(viewport(layout=gl))
pushViewport(vp.1)

par(new=TRUE, fig=gridFIG())
plot.rqss(fit, axes = FALSE, xlab = "", ylab = "",
          render="contour")
popViewport()

pushViewport(vp.2)
ggplotted <- ggplot(house) + geom_point(
  mapping=aes(x=longitude, y=latitude,
              col=median_house_value)) +
  scale_color_gradientn("median\nhouse value",
                        colours = rainbow(7)) +
  theme(legend.position = c(.8, .8))
print(ggplotted, newpage = FALSE)

popViewport(1)

```

##

```{r echo=F,warning=FALSE, message=FALSE, size="small"}

plot.new() 

gl <- grid.layout(nrow=1, ncol=2)
vp.1 <- viewport(layout.pos.col=1, layout.pos.row=1) 
vp.2 <- viewport(layout.pos.col=2, layout.pos.row=1) 

pushViewport(viewport(layout=gl))
pushViewport(vp.1)

par(new=TRUE, fig=gridFIG())
plot.rqss(fit, axes = FALSE, xlab = "", ylab = "",
          render="contour")
popViewport()

pushViewport(vp.2)
ggplotted <- ggplot(house) + geom_point(
  mapping=aes(x=longitude, y=latitude,
              col=median_house_value)) +
  scale_color_gradientn("median\nhouse value",
                        colours = rainbow(7)) +
  theme(legend.position = c(.8, .8))
print(ggplotted, newpage = FALSE)

popViewport(1)

```


```{r echo=FALSE, eval=FALSE, warning=FALSE, message=FALSE, fig.height=6, size="small"}

par(mfrow=c(1,2))
plot(house$longitude,house$latitude)
plot.rqss(fit, axes = FALSE, xlab = "", ylab = "",
          render="contour")
```


